BOLD Parameter Estimation Using Particle Filters

### 2010-07-02 doc/thesis/intro.tex
almost done with methods

\begin{figure}
\centering
\subfigure[]{\label{fig:QQDDlta:A}\includegraphics[trim=6cm 1cm 6cm 0,width=14cm]{images/noise2_0009d_29_49_9}}
\subfigure[]{\label{fig:QQDDlta:B}\includegraphics[trim=6cm 1cm 6cm 0,width=14cm]{images/noise2_0009d_34_43_24}}
\subfigure[]{\label{fig:QQDelta:C}\includegraphics[trim=6cm 1cm 6cm 0,width=14cm]{images/noise2_0009d_22_38_23}}
\subfigure[]{\label{fig:QQDDlta:D}\includegraphics[trim=6cm 1cm 6cm 0,width=14cm]{images/noise2_0009d_37_29_24}}
\caption{Q-Q Plots of resting state data, using the BOLD signal changes}
\label{fig:QQDelta}
\end{figure}

@@

\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c |}
\hline 
Parameter  & \cite{Friston2002} & \cite{Johnston2008} & \cite{Vakorin2007} & \cite{Deneux2006}\\
\hline
$\epsilon$ &  $N(.54 , .1 ^2)$  & $.069 \pm .014 $ & (NC) & .89\\
$\tau_s  $ &  $N(1.54, .25^2)$  & $4.98 \pm 1.07 $ & $2.2$ & 2.04 \\
$\tau_f  $ &  $N(2.46, .25^2)$  & $8.31 \pm 1.51 $ & $.45$ & 5.26\\
$\tau_0  $ &  $N(.98 , .25^2)$  & $8.38 \pm 1.5  $ & $.94$ & .27\\
$\alpha  $ &  $N(.33 , .45^2)$  & $.189 \pm .004 $ & $.4$ (NC) & .63 \\
$E_0     $ &  $N(.34 , .1 ^2)$  & $.635 \pm .072 $ & $.6$ (NC) & .33\\
$V_0     $ &  $.03$ (NC)        & $.0149 \pm .006$ & (NC) & .16\\
\hline
\end{tabular}
\caption{Parameters found by various studies. (NC) indicates that the value
wasn't calculated. \cite{Vakorin2007} made use of the values from \cite{Friston2002}
where not explicitly stated}
\label{tab:Params} 
\end{table}

>>>>

\begin{figure}
\centering
\subfigure[]{\label{fig:QQDDelta:A}\includegraphics[trim=6cm 1cm 6cm 0,width=14cm]{images/noise2_0009d_29_49_9}}
\subfigure[]{\label{fig:QQDDelta:B}\includegraphics[trim=6cm 1cm 6cm 0,width=14cm]{images/noise2_0009d_34_43_24}}
\subfigure[]{\label{fig:QQDDelta:C}\includegraphics[trim=6cm 1cm 6cm 0,width=14cm]{images/noise2_0009d_22_38_23}}
\subfigure[]{\label{fig:QQDDelta:D}\includegraphics[trim=6cm 1cm 6cm 0,width=14cm]{images/noise2_0009d_37_29_24}}
\caption{Q-Q Plots of resting state data, using the BOLD signal changes}
\label{fig:QQDelta}
\end{figure}

@@

\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c | c|}
\hline 
Parameter  & \cite{Friston2002} & \cite{Johnston2008} & \cite{Vakorin2007} & \cite{Deneux2006}\\
\hline
$\epsilon$ &  $N(.54 , .1 ^2)$  & $.069 \pm .014 $ & (NC) & .89\\
$\tau_s  $ &  $N(1.54, .25^2)$  & $4.98 \pm 1.07 $ & $2.2$ & 2.04 \\
$\tau_f  $ &  $N(2.46, .25^2)$  & $8.31 \pm 1.51 $ & $.45$ & 5.26\\
$\tau_0  $ &  $N(.98 , .25^2)$  & $8.38 \pm 1.5  $ & $.94$ & .27\\
$\alpha  $ &  $N(.33 , .45^2)$  & $.189 \pm .004 $ & $.4$ (NC) & .63 \\
$E_0     $ &  $N(.34 , .1 ^2)$  & $.635 \pm .072 $ & $.6$ (NC) & .33\\
$V_0     $ &  $.03$ (NC)        & $.0149 \pm .006$ & (NC) & .16\\
\hline
\end{tabular}
\caption{Parameters found by various studies. (NC) indicates that the value
wasn't calculated. \cite{Vakorin2007} made use of the values from \cite{Friston2002}
where not explicitly stated}
\label{tab:Params} 
\end{table}

### 2010-07-02 doc/thesis/methods.tex
almost done with methods

\begin{table}[t]
\centering
\begin{tabular}{|c || c | c |}
\hline 
Parameter & Distribution & $\mu$ & $\sigma$ \\
\hline
$\tau_0$ & Gamma & .98 & .5  \\
$\alpha$ & Gamma & .33 & .045\\
$E_0$    & Gamma & .34 & .03  \\
$V_0$    & Gamma & .03 & .006 \\
$\tau_s$ & Gamma & 1.54  & .5 \\
$\tau_f$ & Gamma & 2.46  & .5 \\
$\epsilon$ & Gamma & .64 & .4 \\
\hline
\end{tabular}
\caption{}
\label{tab:Prior} 
\end{table}

@@

\section{Choosing ${Pr}(y(T) | x(T))$}
Choosing a representation of an unknown distribution is certainly tricky,
and so the fact that $\hat{Pr}(y(T) | x(T)) = \nu_d + \nu_y$ means that
there is a significant piece of the algorithm that is based primarily conjecture.
Studies of the noise in FMRI typically attribute noise to a Gaussian random
variable or an additive noise process with Gaussian steps.

Another natural choice might be one of the robust estimator weight functions, for
example the Huber or bi-square. For the purpose of this work we will stick
with long tailed distributions, however it is worth noting that long tails
may not be the optimal choice in all, or even this situation. The justification
for long tailed distributions is that we believe the noise to be long tailed,
and the variance of the noise is not well known.

Therefore, we tried three weighting functions based on three distributions: Gaussian, 
Laplace and the Cauchy. The standard deviation of the distribution is extremely
important to the convergence of particle filter. A standard deviation that is 
too large will not allow the distribution to converge in any reasonable number of 
measurements. A standard deviation below the standard deviation of the noise 
will cause the algorithm to throw out perfectly acceptable particles. 
The weighting function ultimately will shape the output distribution, $P[y]$, into that
distribution, however the distribution of $x$ will still approach a reasonable
estimate of its true distribution. Even an overly wide weighting function, will
allow the Gaussian Mixture estimate of $X$ to converge to the correct location 
parameters of the "real" posterior distribution, though the scale parameters may 
be overly large.

A reasonable method of setting the standard deviation of $\Omega$ may be by 
taking a small sample from "resting" data and using the sample standard deviation.
Since this is the first attempt at using particle filters for modeling the 
BOLD model, in this work we set the standard deviation manually at <weight standard dev>,
because it gives a more consistency and control. Of course this could be taken
further, by testing the sample data against a set of stock distributions
and choosing the best fit. This of course depends on having enough samples
to make a reasonable inference, which may not always exist.

\subsection{Classical De-trending}
The non-stationary
aspect of a Weiner process as with $\nu_d$ is difficult to compensate for, and so various methods
have been developed to compensate for it. \cite{Tanabe2002} and \cite{Smith1999} have
demonstrated that this component is prevalent, and may in fact be a characteristic
of FMRI. In some studies, as many as half the voxels benefit from detrending, meaning
that this is certainly a serious barrier to inference. All the existing methods are performed
during the preprocessing stage, rather than as an integral part of analyzing the BOLD
signal. There is no shortage of theories on the "best" method of detrending; however
a head to head comparison, \cite{Tanabe2002}, showed that in most cases subtracting off
a spline works the best. The benefit of the spline versus wavelets, high pass 
filtering or other DC removal techniques is that the frequency response is not set.
Rather, the spline is adaptive to the input, having a low cut off if the signal's 
median stays constant but a high cut off frequency if the signal's median tends to
shift heavily over the course of time. In spite of this, the spline will still remove some
amount of signal, just like all of these methods, which is why the the method proposed
in \autoref{sec:Methods Delta Based Inference} was considered.

When using the median-based spline techniques, the resulting signal is will almost
certainly end with a median of zero; in other words half the signal will be above zero,
half below. The problem with this is this is not the "natural" BOLD signal. More specifically,
when the signal is inactive, the BOLD response \emph{should} be at 0\% change from
the base level; activation may then increase, or for short periods decrease from this base.
After removing the spline, the BOLD resting state will be below 0\%.
This is troublesome because it reduces the power the BOLD model. One obvious solution
is to add an arbitrary constant to each BOLD response. Of course this won't scale well
to whole brain analysis. A more effective technique is adding a DC gain to the BOLD
model. Essentially every measurement would then have a constant level added to it. 
Like all the other model parameters, enough measurements should allow the correct
value to be found. On the downside adding another dimension increases the
complexity of the model, for a variable that is relatively obvious to a human.

Thus, a more conventional approach was used. A robust estimator of scale, was used
to determine roughly where the base level was. The Median Absolute Deviation (MAD)
proved to be extremely accurate in determining how much to shift the signal up
by. We tested both methods during the course of analysis, and found that the increase 
in model complexity far outweighted the slight increase in flexibility. Its 
possible that a more accurate method may exist; however, for this case the 
MAD seemed to work perfectly, as \autoref{fig:DesplineQuality} shows.

@@

\subsection{Delta Based Inference}
\label{sec:Methods Delta Based Inference}
I also propose and test a different method of dealing with the so called "drift". 
Instead of comparing the direct output of the particle filter with the direct
measurement, the algorith compares the change in signal over a single TR. 
In most signal processing cases this would foolish, but that is because the 
general assumption that all noise is high frequency is not the case here.
In fact, every pipeline for the analysis of BOLD signal uses a high pass filter,
but low poss filters are rarely applied, because it is a well known fact that 
most of the signal is in the high frequency range and most of the noise is actually 
in the low frequency range. The particle filter is an 
extremely robust method of inference, and so I would assert that the particle
filter ought to be given as \emph{raw} data as possible. While taking direct measurements
without de-trending would give awful results, using the difference removes the 
DC component and turns a Weiner process into a Gaussian random variable.

@@

Because $\nu_d$ is a Weiner process, then $\nu_d(t) - \nu_d(t-1)$ is simply a Gaussian step. If
$\nu_d$ is some other additive process, the difference will still be one of a few stable
distributions. If $\nu_y$ is i.i.d. then the resulting distribution will still be zero mean
with a maximum variance of twice the original variance. All the assumptions made originally
for the particle filter hold, and all of the parameters may be distinguished based on
the step sizes, thus it is not unreasonable to attempt to match the string of step sizes
rather than string of direct readings.

@@

\section{Preprocessing}
\label{sec:Methods Preprocessing}
As discussed in the section on de-trending, the normal pipeline for analyzing
FMRI involves a great deal of preprocessing. In this paper we make an effort to
minimize any type of preprocessing that will degrade the signal. 
After FMRI data has been acquired it is always necessary to modify the
data in some way to make different runs comparable. Because FMRI signal
levels are unit-less, at the very least it is necessary to convert
the data into \% difference from the baseline. This process removes no data
from signal since it merely subtracting then dividing by a constant. This
is the signal that was input into the delta based particle filter.
Of course there are much more advanced ways of performing this task.
The generally accepted standard is actually to use a high pass filter, although the
cutoff frequency is application dependent and often applied haphazardly.
The high pass filter thus removes the DC component of the signal, and 
some amount of the so called "drift". The problem with this method is that it is
not adaptive to the input. Huge variations in drift frequencies can exist 
in a single time-series. Thus, a single cutoff frequency could miss a significant
drift component, or it could remove \emph{actual} signal, if the cutoff frequency is
set too high. This is why, as I mentioned in the De-trending section, a spline
based detrending method will generally give better results.

>>>>

\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c |}
\hline 
Parameter & Distribution & $\mu$ & $\sigma$ \\
\hline
$\tau_0$ & Gamma & .98 & .5  \\
$\alpha$ & Gamma & .33 & .045\\
$E_0$    & Gamma & .34 & .03  \\
$V_0$    & Gamma & .03 & .006 \\
$\tau_s$ & Gamma & 1.54  & .5 \\
$\tau_f$ & Gamma & 2.46  & .5 \\
$\epsilon$ & Gamma & .64 & .4 \\
\hline
\end{tabular}
\caption{}
\label{tab:Prior} 
\end{table}

@@

\section{Choosing $P(y_k | x_k)$}
Choosing a representation of an unknown distribution tricky, and while
the usual choice is a Gaussian distribution, there are reasons why it
may not be the best.  Studies of the noise in FMRI typically characterise
noise as an additive noise process with Gaussian steps; a Wiener process.

As noted in \autoref{sec:Introduction Noise}, the noise is not strictly Gaussian,
nor is it strictly Wiener. As with any unknown noise however, it is necessary 
to make some assumption. If the weighting function ($P(y_k | x_k)$) exactly
matches the measurement error, then the ideal particle filter will result.
Particles with $x_k$'s that estimate $y_k$ far out on the weighting
function will quickly have weights near 0. Thus, an weighting function that
exactly matches $P(Y(t) | X(t))$ will easily, and correctly throw out incorrect
particles. However assuming that this function is not going to be found 
for arbitrary voxels, it is necessary to choose a function to approximate this 
probability. The cost of choosing an overly broad distribution for this
function is an overly broad $P(x_k | y_k)$; and ultimately an overly
broad $P(x_k | y_{0:k})$. On the other hand, an overly thin distribution
will lead to, at best, an overly thin representation of these same 
distributions; but at wost will lead to particle deprivation (all particles
being zero-weighted). If the \emph{true} distribution were found, particle
deprivation would indicate that that time series is not being affected
in any way describable by the prior distribution/model. Particle deprivation
due to an under-variant weighting function gives no information. Thus,
the cost of under variance is significantly higher; a fact that definitely needs
to be taken into account.

Because of the need to err on the side of caution I tested several weighting
functions. In addition to the Gaussian I also tested the Laplace and Cauchy
distributions, both of which have much wider tales than the Gaussian. The
benefit of the wider tailed distributions is that they don't down-weight
particles quite as fast. Additionally, the Laplace distribution has the
benefit of having a non-zero slope at the origin. This means that even
if the distribution is made overly broad, it will still distinguish between
particles that are near the origin. Results with real data compare 
the three distribution in \autoref{sec:Results Weights}.

After trial and error, I chose the weight standard deviation to be $.005$. Obviously
the case of the Cauchy, this is simply the scale parameter. While I 
did attempt to automatically set the standard deviation, I found that the algorithm
became too unpredictable. Essentially, if the weight function isn't
fixed across voxels, very noisy time series with no actual signal will 
converge to nonsensical results. 
In the future, it may be possible to set the standard deviation by
taking a small sample from "resting" data and using the sample standard deviation.
Since this is the first attempt at using particle filters for modeling the 
BOLD model, in this work I set the standard deviation manually at $.005$,
because it gives a more consistency and control.

\subsection{Classical De-trending}
The non-stationary
aspect of a Weiner process, presumably the result of integrating some
$\nu_x$ is difficult to compensate for, and so various methods
have been developed to compensate for it. \cite{Tanabe2002} and \cite{Smith1999} have
demonstrated that this component is prevalent, and may in fact be an inherent  characteristic
of FMRI. In some studies, as many as half the voxels benefit from detrending, presenting
a serious barrier to inference (\cite{Smith2007}). All the existing methods are performed
during the preprocessing stage, rather than as an integral part of analyzing the BOLD
signal. There is no shortage of theories on the "best" method of detrending; however
in a head to head comparison, \cite{Tanabe2002}, showed that in most cases subtracting off
a spline works the best. The benefit of the spline versus wavelets, high pass 
filtering or other DC removal techniques is that the frequency response is not set.
Rather, the spline is adaptive to the input, having a low cut off if the signal's 
median stays constant but a high cut off frequency if the signal's median tends to
shift heavily over the course of time. In spite of this, the spline will still remove some
amount of signal, just like all of these methods, which is why I considered the method proposed
in \autoref{sec:Methods Delta Based Inference}.

The method I used to calculate the spline was picking one knot for every 15
measurements in an image. Thus a 10 minute session at a repetition time of 
2.1 seconds would have 19 knots. The knot first and last knots were each 
given half the number of samples as the rest of the knots; which were all 
located at the center of their sample group. The median of each sample group
was then taken and used as the magnitude for the group. Taking the median 
versus the mean seemed to work better, given the presence of outliers. It seems
that there is potential to optimize the spline further using a canonical 
HRF to find resting points; however, for this to work the experiment would have
to be designed with this in mind.

When using the median-based spline techniques, the normalized signal will,
by definition have a median near zero. 
The problem with this is this is not the "natural" BOLD signal. More specifically,
when the signal is inactive, the BOLD response \emph{should} be at 0\% change from
the base level; activation may then increase, or for short periods decrease from this base.
After removing the spline, the BOLD resting state will be below 0\%.
This is problematic because it reduces the ability of an algorithm to learn.
One quick and dirty solution is to add an arbitrary constant to each BOLD response. 
Of course this won't scale to whole-brain analysis, so a more effective technique 
is adding a DC gain parameter to the BOLD model. 
Like all the other model parameters, given enough measurements, correct
value may be found. On the downside, adding another dimension increases the
complexity of the model, for a variable that is relatively obvious by direct
observation.

Thus, I used a more conventional approach to deal with this. To determine
the DC gain to be added to the signal I used a robust estimator 
of scale. The Median Absolute Deviation (MAD)
proved to be extremely accurate in determining how much to shift the signal up
by. I tested both methods during the course of analysis, and found that the increase 
in model complexity far outweighed the slight increase in flexibility. Its 
possible that a more accurate method may exist; however, for this case the 
MAD seems to work as it should, as \autoref{fig:DesplineQuality} shows.

\begin{equation}
y_{\text{gain}, 0:K} = 2\text{median}_{i=0:K}(y_i - \text{median}(y_{0:K}))
\end{equation}

@@

A serious concern with adding and subtracting arbitrary values to 
real data is whether this will create false positives. This is a legitimate
concern; however, all the additions and subtractions done in this section
have been very low frequency changes and should not substantially change
the BOLD response, being a relatively high frequency signal. The typical
method of preprocessing is a high-pass filter with a cutoff of around 30
seconds; which should about match the spline with knots every fifteen 
measurements. \cite{Tanabe2002} found that splines tended to far outperform the
high-pass filter method.

\subsection{Delta Based Inference}
\label{sec:Methods Delta Based Inference}
The alternative to these sorts of low frequency manipulation is to
go around the problem in another way. Here, I propose and test a 
different method of dealing with the so called "drift". 
Instead of comparing the direct output of the particle filter with the direct
measurement, the algorithm would compare the change in signal over a single TR,
with the result of integrating the model for the same period. 
In most signal processing cases this would be foolish, but that is because the 
general assumption is that all noise is high frequency. Considering 
the fact that every BOLD analysis pipeline uses a high pass filter,
whereas low poss temporal filter are rarely applied, it makes sense
that a derivative type method could work. The benefit of particle filters
is that they are extremely robust method of inference, and I would assert 
that the particle filter ought to be given as \emph{raw} data as possible. 
While taking direct measurements
without de-trending would give awful results, using the difference removes the 
DC component and turns what is usually assumed to be a Weiner process into 
a simple Gaussian random variable.

@@

Even if $\nu_d$ is some other additive process, the difference will still be closer
to I.I.D. than a Wiener process, as the autocorrelation of the $\delta y$ shows
in \autoref{fig:QQDDelta} in \autoref{sec:Introduction Noise}. 
 All the assumptions made originally
for the particle filter still hold, and all of the parameters may be distinguished based on
the step sizes, thus it is not unreasonable to consider matching the string of step sizes
rather than string of direct readings.

@@

\section{Preprocessing}
\label{sec:Methods Preprocessing}
As discussed in the section on de-trending, the normal pipeline for analyzing
FMRI involves a great deal of preprocessing. In this paper we make an effort to
minimize any type of preprocessing that will modify the signal. 
After FMRI data has been acquired it is always necessary to modify the
data in some way to make different runs comparable. Because FMRI signal
levels are unit-less, at the very least it is necessary to convert
the data into \% difference from the baseline. This process removes no data
from signal since it merely subtracting then dividing by a constant. This
is the signal that was input into the delta based particle filter.
Of course there are much more advanced ways of performing this task.
The generally accepted standard is actually to use a high pass filter, although the
cutoff frequency is application dependent and often applied haphazardly.
The high pass filter thus removes the DC component of the signal, and 
some amount of the so called "drift". The problem with this method is that it is
not adaptive to the input. Huge variations in drift frequencies can exist 
in a single time-series. Thus, a single cutoff frequency could miss a significant
drift component, or it could remove \emph{actual} signal, if the cutoff frequency is
set too high. This is why, as I mentioned in the De-trending section, a spline
based detrending method will generally give better results.
